Technical comment on 'Negative-assortative mating for color in wolves

### 2022-08-17 ms-r2/ms.Rmd
v2.1

<!-- \begin{centering} -->
<!-- Christopher D. Muir\textsuperscript{1,*} -->

<!-- \bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA -->
<!-- \end{centering} -->

@@

> "gray wolves have a genotype of $kk$ and an assumed frequency of $P$ and black
wolves have genotypes $Kk$ and $KK$ with frequencies $H$ and $Q$, respectively $(P + H + Q = 1)$."

@@

1. $kk$, $Kk$, and $KK$ are mutually exclusive genotypes with frequencies $P$, $H$, and $Q$
2. Negative-assortative mating and random mating are mutually exclusive mating types with frequencies $A$ and $1-A$
3. Genotype and mating type are independent ($\mathrm{Pr}[A \cap B] = \mathrm{Pr}[A] \times \mathrm{Pr}[B]$)

Other standard population genetic assumptions such as infinite population size also apply.

@@

To summarize, a proportion $A$ should mate assortatively given the assumptions of their model, but only $2 AP(H+Q)$ actually mate assortatively according to their results. In essence, they assign a proportion $A$ to mate assortatively, but then a proportion $A - 2 AP(H+Q)$ do not mate assortatively (the area of the gray regions in Fig. \@ref(fig:sample-space)b), and are therefore not counted among the total number of matings. This is why the probabilities of all matings do not sum to 1. The logically consistent solution is to condition on the fact that if a mating is negative-assortative it must by definition have one gray and one black parent (Fig. \@ref(fig:sample-space)b). In the next section, I use this approach to derive new results. In contrast, @hedrick_negative-assortative_2016 effectively impose selection without ever stating that assumption. To deal with the fact that resulting genotype frequencies do not sum to 1, they regularize the frequencies (*cf* equation 1a-b), as normally done in models of selection. Regularization is appropriate with selection because selection shrinks or expands the sample space as long as average fitness does not equal 1. @hedrick_negative-assortative_2016 impose selection because some of the assortative-mating individuals do not mate assortatively and are therefore not counted among the matings that result in offspring.

@@

The previous section showed that the frequency of assortative gray $\times$ black matings was not derived in a manner logically consistent with the model's assumptions. Here I derive new mating frequencies, genotype frequencies, and equilibria. I used Sympy version `r sympy_version` [@meurer_sympy:_2017] for symbolic derivations through Python version `r py_info[["version"]]` and the *R* package **reticulate** version `r pkg_info["reticulate"]` [@ushey_reticulate_2022]. All other computations were performed in *R* version `r getRversion()` [@r_core_team_r:_2022]. The source code is available in a public GitHub repository and will be archived on Zenodo upon publication.
<!-- The source code is available in a public GitHub repository (https://github.com/cdmuir/disassortative-mating) and will be archived on Zenodo upon publication. -->

Table \@ref(tab:probabilities) derives the probabilities of all possible outcomes and Table \@ref(tab:genotypes) summarizes the frequency each mating combination. This is the exact same process used to derive frequency of mating combinations in positive-assortative mating models [e.g. @hedrick_population_2012]. Since I model random mating identically to the previous model, the frequencies of gray $\times$ gray and black $\times$ black matings are identical; only the frequency of gray $\times$ black matings differs between models (Table \@ref(tab:genotypes); Fig. \@ref{fig:sample-space}c).

@@

\begin{align*}
  \mathrm{Pr}[\textrm{gray} \times \textrm{black}] = & 1 \times A + 2P(H+Q) \times (1-A) \\
  = & A + 2 P (H + Q) (1 - A)
\end{align*}

This result diverges from that given in @hedrick_negative-assortative_2016, where they report the frequency of gray $\times$ black matings is $2 P (H + Q)$ (*cf* Table 1).

>>>>

\begin{centering}
Christopher D. Muir\textsuperscript{1,*}

\bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA
\end{centering}

@@

> "gray wolves have a genotype of $\mathit{kk}$ and an assumed frequency of $P$ and black
wolves have genotypes $\mathit{Kk}$ and $\mathit{KK}$ with frequencies $H$ and $Q$, respectively $(P + H + Q = 1)$."

@@

1. $\mathit{kk}$, $\mathit{Kk}$, and $\mathit{KK}$ are mutually exclusive genotypes with frequencies $P$, $H$, and $Q$
2. Negative-assortative mating and random mating are mutually exclusive mating types with frequencies $A$ and $1-A$
3. Genotype and mating type are independent ($\mathrm{Pr}[A \cap B] = \mathrm{Pr}[A] \times \mathrm{Pr}[B]$)

Other model assumptions made by @hedrick_negative-assortative_2016 are common in population genetics. The model is deterministic, individuals are monoecious, generations are discrete and nonoverlapping, and individuals can mate multiply. Some of these assumptions are not be valid for wolves and different assumptions could result in qualitatively different outcomes. For example, if individuals only mate in pairs, this would constrain the amount of negative assortative mating that is possible when either allele is rare. While alternative assumptions may help understand how assortative mating maintains genetic variation, my purpose here is not to critique the assumptions of @hedrick_negative-assortative_2016, but rather to show their results are not logically consistent with their stated assumptions.

@@

To summarize, a proportion $A$ should mate assortatively given the assumptions of their model, but only $2 AP(H+Q)$ actually mate assortatively according to their results. In essence, they assign a proportion $A$ to mate assortatively, but then a proportion $A - 2 AP(H+Q)$ do not mate assortatively (the area of the gray regions in Fig. \@ref(fig:sample-space)b), and are therefore not counted among the total number of matings. This is why the probabilities of all matings do not sum to 1. The logically consistent solution is to condition on the fact that if a mating is negative-assortative it must by definition have one gray and one black parent (Fig. \@ref(fig:sample-space)b). In the next section, I use this approach to derive different results. In contrast, @hedrick_negative-assortative_2016 effectively impose selection without ever stating that assumption. To deal with the fact that resulting genotype frequencies do not sum to 1, they regularize the frequencies (*cf* equation 1a-b), as normally done in models of selection. Regularization is appropriate with selection because selection shrinks or expands the sample space as long as average fitness does not equal 1. @hedrick_negative-assortative_2016 impose selection because some of the assortative-mating individuals do not mate assortatively and are therefore not counted among the matings that result in offspring.

@@

The previous section showed that the frequency of assortative gray $\times$ black matings was not derived in a manner logically consistent with the model's assumptions. Here I derive new mating frequencies, genotype frequencies, and equilibria. This model was previously analyzed by @li_first_1976] where in his notation $D = H + Q$, $R = P$, and $w = A$ (*cf* pg. 367-368). I used Sympy version `r sympy_version` [@meurer_sympy:_2017] for symbolic derivations through Python version `r py_info[["version"]]` and the *R* package **reticulate** version `r pkg_info["reticulate"]` [@ushey_reticulate_2022]. All other computations were performed in *R* version `r getRversion()` [@r_core_team_r_2022]. The source code is available in a public GitHub repository and will be archived on Zenodo upon publication.
<!-- The source code is available in a public GitHub repository (https://github.com/cdmuir/disassortative-mating) and will be archived on Zenodo upon publication. -->

Table \@ref(tab:probabilities) derives the probabilities of all possible outcomes and Table \@ref(tab:genotypes) summarizes the frequency each mating combination. This is the exact same process used to derive frequency of mating combinations in positive-assortative mating models [e.g. @hedrick_population_2012]. Since I model random mating identically to the previous model, the frequencies of gray $\times$ gray and black $\times$ black matings are identical; only the frequency of gray $\times$ black matings differs between models (Table \@ref(tab:genotypes); Fig. \@ref(fig:sample-space)c).

@@

\begin{align}
  \mathrm{Pr}[\textrm{gray} \times \textrm{black}] = & 1 \times A + 2P(H+Q) \times (1-A) \nonumber \\
  = & A + 2 P (H + Q) (1 - A) \label{eq1}
\end{align}

This result diverges from that given in @hedrick_negative-assortative_2016, where they report the frequency of gray $\times$ black matings is $2 P (H + Q)$ (*cf* Table 1). One advantage of their result is that the frequency of assortative matings is 0 when the population is fixed at this locus ($P = 0$ or $H + Q = 0$). In contrast, Eq. \ref{eq1} conjures $A$ assortative matings from a population of wolves with identical coat color. This is a logical consequence of assuming that a proportion $A$ mate assortatively irrespective of the allele frequency. An alternative approach is to set bounds on $A$ as a function color morph frequency. For example, if we assume that wolves mate in monogamous pairs, then $A$ cannot exceed twice the frequency of the rarer color morph. Consider a single black wolf in pack of 100 ($P = 0.99$). The maximum value of $A$ is $1/50$ if there is one gray $times$ black mating and 49 gray $times$ gray matings. More generally, the maximum value of $A$ in a population of monogamous pairs can be found by solving Equation \ref{eq1} for $A$ when the maximum proportion of assortative matings is the lesser of $2P$ and $2(1 - P)$. When grey morphs are rarer than black morphs ($P < 0.5$), the maximum value of $A$ is $2P ^ 2 / (2P ^ 2 - 2 P + 1)$; when black morphs are rarer ($P > 0.5$), the maximum value of $A$ is $2 (P^2 - 2P + 1) / (2 P ^ 2 - 2 P + 1)$. Different constraints on $A$ apply to a population where individuals can self-fertilize and/or mate with multiple individuals. Incorporating constraints on $A$ as a function of phenotypic frequency may be important, but the assumptions stated by @hedrick_negative-assortative_2016 do not imply constraint on $A$ when one color morph is rare.
